---
title: "Mixing_Models"
author: "Tali Caspi"
date: "2025-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install MixSIAR if needed
# remotes::install_github("brianstock/MixSIAR", dependencies=T, version="3.1.12")

# load packages
library(tidyverse) # version 2.0.0
library(MixSIAR) # version 3.1.12
library(plotrix)
library(parallel)

# Custom theme for plotting
theme_custom <- function() {
  theme_classic()+
  theme(panel.grid.minor = element_blank(),
        strip.text = element_text(size = 12, face = "bold"),
        plot.background = element_rect(fill = "white", color = NA),
        plot.title = element_text(size = 7),
        strip.background = element_blank(),
        axis.text.y = element_text(size=10, color="black"),
        axis.text.x = element_text(size=10, color="black"),
        axis.title.y = element_text(size=10, color="black"),
        axis.title.x = element_text(size=10, color="black"))}

# Set colors
cols <- c("ff_chicken" = "#C0362D",
          "wet.food"= "#F57E77",
          "medium.mammal"= "#687E90", 
          "rat"= "#B6CEE2",
          "gopher"= "#004D40",
          "fruit"= "#59BBAA")

cols.paired <- c("Mean_ff_chicken" = "#C0362D",
          "Mean_wet.food"= "#F57E77",
          "Mean_medium.mammal"= "#687E90", 
          "Mean_rat"= "#B6CEE2",
          "Mean_gopher"= "#004D40",
          "Mean_fruit"= "#59BBAA")
```

These MixSIAR models take a long time to run; it is recommended to run them on an HPC cluster.

# Model with categorical effects and six dietary sources

## Prepare source data

```{r}
# Load fast food data
ff <- read.csv("Data/fastfood_si.csv")

# Load prey data
prey <- read.csv("Data/preydata_si.csv")

# define outliers and sources to keep
remove.sources <- c("Goph10") # sample removed because it was a strong outlier
select.sources <- c("gopher", "fruit", "skunk", "raccoon", "cat", "rat", "wet.food")

# Step 1: Compute mean and SD for each species
species_stats <- prey %>%
  
  # remove outliers
  filter(!SampleID %in% remove.sources) %>%
  
  # select target species
  filter(species %in% select.sources) %>% 
  
  # adjust hair back to muscle
  mutate(d13C = if_else(material %in% c("vertebrate hair"), d13C - 1.31, d13C)) %>% 
  
  # select cols
  dplyr::select(c(SampleID, species, d13C, d15N)) %>% 
  
  # rbind chicken data
  rbind(ff %>% filter(species == "ff_chicken")) %>% 
  
  # summarize
  group_by(species) %>%
  summarise(
    Meand13C = mean(d13C),
    SDd13C = sd(d13C),
    Meand15N = mean(d15N),
    SDd15N = sd(d15N),
    n = n())

# Step 2: Compute grand mean and pooled SD for "medium.mammal"
medium_mammal_stats <- species_stats %>%
  filter(species %in% c("raccoon", "skunk", "cat")) %>%
  summarise(
    species = "medium.mammal",
    Meand13C = mean(Meand13C),  
    Meand15N = mean(Meand15N),

    # Pooled standard deviation formula
    SDd13C = sqrt(sum((SDd13C^2 * (n - 1))) / sum(n - 1)),
    SDd15N = sqrt(sum((SDd15N^2 * (n - 1))) / sum(n - 1)),
    n = sum(n))

# Step 3: Filter only the desired species and add "medium.mammal"
source_df <- species_stats %>%
  filter(species %in% c("gopher", "fruit", "rat", "wet.food", "ff_chicken")) %>%
  bind_rows(medium_mammal_stats)

# Write to csv
# write.csv(source_df, "Cleaned_Data/coyote_source.csv", row.names = FALSE)
```

## Prepare TDF data frame

I'm pulling values from Ryan Stephen's paper: https://esajournals.onlinelibrary.wiley.com/doi/10.1002/ecm.1525

```{r}
# Create TDF dataframe
tdf_df <- data.frame(
  species = source_df$species,
  Meand13C = 3.97,
  SDd13C = 1.85,
  Meand15N = 2.76,
  SDd15N = 1.41)

# write.csv(tdf_df, "Cleaned_Data/coyote_discrimination.csv", row.names = FALSE)
```

## Prepare consumer data

```{r}
# Load consumer data 
data_averages <- read.csv("Cleaned_Data/averaged_whiskerdata.csv") %>% 
  filter(urban == "urban") %>%  # keep only SF coyotes
  mutate(group = ifelse(is.na(group), "Unknown", group)) #rename NA to Unknown for territories

# Make consumer df
consumer_df_avgs <- data.frame(
  d13C = data_averages$C_avg,
  d15N = data_averages$N_avg,
  individual = data_averages$whisker,
  group = data_averages$group) %>% 
  mutate_if(is.character, as.factor) 

# Write to file
# write.csv(consumer_df_avgs, "Cleaned_Data//coyote_consumer_avgs.csv", row.names = FALSE)
```

## Investigate mixing polygon

```{r}
# set TDFs
TDF.C <- 3.97
TDF.N <- 2.76

# Make data frame for plotting with source values corrected by TDF factors
source_df_plotting <- source_df %>% 
  mutate(Meand13C = Meand13C + TDF.C,
         Meand15N = Meand15N + TDF.N)

# Data frame for my chicken/pork samples
meat.names <- c("ChkC01", "Pork01")

meat_samples <- prey %>% 
  filter(SampleID %in% meat.names) %>% 
  dplyr::select(SampleID, d13C, d15N) %>% 
  mutate(d13C = d13C + TDF.C,
         d15N = d15N + TDF.N)

# Plot prey sources
fig.4a <- ggplot()+
  geom_point(data=source_df_plotting, mapping=aes(x=Meand13C, y=Meand15N, color=species),
             size=3)+
  geom_errorbar(data=source_df_plotting, mapping=aes(x=Meand13C, y=Meand15N,
                                                 ymax=Meand15N+SDd15N, ymin=Meand15N-SDd15N,
                                                 width=0.3, color=species))+
  geom_errorbar(data=source_df_plotting, mapping=aes(x=Meand13C, y=Meand15N,
                                                 xmax=Meand13C+SDd13C, xmin=Meand13C-SDd13C,
                                                 width=0.3, color=species))+
  geom_point(data=consumer_df_avgs, mapping=aes(x=d13C, y=d15N), size=1, alpha=0.7)+
  
  # Meat samples
  geom_point(data=meat_samples, mapping=aes(x=d13C, y=d15N), alpha=0.9, fill="#C0362D", shape=23, color="black")+
  
  theme_custom()+
  scale_y_continuous(limits=c(4,13), breaks = seq(4, 15, by = 1))+
  scale_x_continuous(limits=c(-25,-11.3), breaks = seq(-25, -10, by=1))+
  xlab(expression(delta^{13}~C)) +
  ylab(expression(delta^{15}~N))+
  theme(legend.position = "none")+
  scale_color_manual(values=cols)

fig.4a
```

## Create MixSIAR model objects

```{r}
mix.coy.avgs <- load_mix_data(filename="Cleaned_Data/coyote_consumer_avgs.csv", 
                     iso_names=c("d13C","d15N"), 
                     factors=c("individual","group"), 
                     fac_random=c(TRUE, TRUE), 
                     fac_nested=c(TRUE, FALSE), 
                     cont_effects=NULL)

source.coy <- load_source_data(filename="Cleaned_Data/coyote_source.csv",
                           source_factors=NULL, 
                           conc_dep=FALSE, 
                           data_type="means", 
                           mix.coy.avgs)

discr.coy <- load_discr_data(filename="Cleaned_Data/coyote_discrimination.csv", mix.coy.avgs)
```

## Set prior

Run the model with an informative prior set based on percet of occurrence for each dietary source for DNA metabarcoding data (see Caspi et al. 2025).

Chicken: 21.5%
Fruit: 20.2%
Gopher: 20.3%
Medium Mammals: 8.5%
Rats: 8.0%
Wet food: 21.5%

```{r}
# Set prior (alphabetical order for sources)
coy.alpha <- c(21.5, 20.2, 20.3, 8.5, 8.0, 21.5)
coy.alpha <- coy.alpha*length(coy.alpha)/sum(coy.alpha)

plot_prior(alpha.prior=coy.alpha,
           source=source.coy,
           plot_save_pdf=FALSE,
           plot_save_png=FALSE,
           filename="prior_plot_coy")
```

## Run model

Write the JAGS model file and run it. This takes a very long time (~ 24 hours), so it is recommended to run this model on an HPC cluster.

```{r}
# Write the JAGS model file
model_filename <- "MixSIAR_model_full.txt"
resid_err <- TRUE
process_err <- TRUE
write_JAGS_model(model_filename, resid_err, process_err, mix.coy.avgs, source.coy)
 
# Run full model - best run on HPC, takes a long time to run
jags.full <- run_model(run="very long", mix.coy.avgs, source.coy, discr.coy, model_filename, alpha.prior=coy.alpha)

# Load model output
jags.full <- readRDS("Model_Output/jags.full.rds")

# Set model output options
output_options <- list(summary_save = TRUE,                 
                       summary_name = "summary_statistics", 
                       sup_post = TRUE,                    
                       plot_post_save_pdf = FALSE,           
                       plot_post_name = "posterior_density",
                       sup_pairs = TRUE,             
                       plot_pairs_save_pdf = FALSE,    
                       plot_pairs_name = "pairs_plot",
                       sup_xy = TRUE,           
                       plot_xy_save_pdf = FALSE,
                       plot_xy_name = "xy_plot",
                       gelman = TRUE,
                       heidel = FALSE,  
                       geweke = TRUE,   
                       diag_save = FALSE,
                       diag_name = "diagnostics",
                       indiv_effect = FALSE,       
                       plot_post_save_png = FALSE, 
                       plot_pairs_save_png = FALSE,
                       plot_xy_save_png = FALSE,
                       diag_save_ggmcmc = TRUE,
                       return_obj = TRUE)

# Assess convergence with Gelman-Rubin and Geweke dignostics
output_diagnostics(jags.full, mix.coy.avgs, source.coy, output_options)

# Mean posterior estimates from model
df.stats <- output_stats(jags.full, mix.coy.avgs, source.coy, output_options)

# Global proportion df
df.stats[grep("^p\\.global", rownames(df.stats)), ] %>% round(3) * 100

# Extract posterior distributions for global proportions
global.prop <- as.data.frame(jags.full$BUGSoutput$sims.list$p.global)
names(global.prop) <- c("ff_chicken", "fruit", "gopher", "medium.mammal", "rat", "wet.food")

posterior_global <- global.prop %>% 
  pivot_longer(cols = everything(), names_to = "sources", values_to = "value") %>% 
  mutate(sources = factor(sources, levels = c("ff_chicken", "wet.food", "rat", "medium.mammal", "fruit", "gopher")))
```

## Make plots:

Global diet source proportions:

```{r}
# Plot global proportions
names <- c("Chicken", "Pet Food","Rats", "Medium-Sized \nMammals", "Fruit","Gophers")

ggplot(posterior_global, aes(x = value, fill = sources, color = sources)) +
  geom_density(aes(y = after_stat(scaled)), alpha = 0.3, linewidth=0.8) +
  labs(x = "Proportion",
       y = "Scaled Posterior Density") +
  scale_x_continuous(limits=c(0,1), expand=c(0,0), labels = seq(0, 1, by = 0.25))+
  scale_y_continuous(limits=c(0,1.01), expand=c(0,0), labels = seq(0, 1, by = 0.25))+
  theme_custom() +
  scale_fill_manual(values=cols, labels=names) +
  scale_color_manual(values=cols, labels=names) +
  theme(legend.title = element_blank(),
        legend.key.size = unit(0.3, "cm"),
        legend.position = "bottom",
        legend.justification = "center",
        legend.text = element_text(size=5))+
  guides(fill = guide_legend(nrow = 1))

# ggsave("Figures/FigureS4.png", dpi=600, height=3, width=4)
```

Individual diet source proportions:

```{r}
# Load covariate data
covs <- read.csv("Data/territory_covs.csv")

# Extract individual proportions
ind.props <- as.data.frame(df.stats) %>% 
  rownames_to_column("sample") %>% 
  mutate(rowID = seq(1:nrow(.)), .before=Mean) %>% 
  slice(11:592) %>% 
  separate(sample, into = c("prefix", "whisker", "source"), sep = "\\.", extra = "merge") %>%
  dplyr::select(c(whisker, source, Mean, SD)) %>% 
  pivot_wider(names_from = source, values_from = c(Mean, SD))

# Prep data frame for plotting
ind.props_plotting_all <- ind.props %>% 
  dplyr::select(1 | starts_with("Mean")) %>% 
  pivot_longer(cols = starts_with("Mean"), 
               names_to = "Source", 
               values_to = "Proportion") %>% 
  left_join(data_averages %>% dplyr::select(c(whisker, group, individual, sfcoy)) , by="whisker") %>% 
  group_by(whisker) %>% 
  mutate(Proportion = Proportion / sum(Proportion)) %>% 
  ungroup() %>% 
  left_join(covs, by="group") %>% 
  mutate(group = factor(group, levels = c("Coit", "Bernal", "Corona Heights", "St. Francis", "Glen", "GGP - East", "GGP - West", "McLaren", "Merced", "Presidio", "Unknown"))) %>% 
  mutate(Source = factor(Source, levels = c("Mean_ff_chicken", "Mean_wet.food", "Mean_medium.mammal", "Mean_rat", "Mean_fruit", "Mean_gopher")))

whisker_to_sfcoy <- ind.props_plotting_all %>%
  dplyr::select(whisker, sfcoy) %>%
  distinct() %>%
  deframe()


# source labels
mean.names <- c("Chicken", "Cat Food", "Medium-Sized \nMammals", "Rats", "Fruit", "Gophers")

# plot
fig.4b <- ind.props_plotting_all %>%
    ggplot(aes(y = reorder(whisker, group), x = Proportion, fill = Source)) +
    geom_col() +
    scale_x_continuous(expand = c(0, 0)) +
    facet_grid(group ~ ., scales = "free", space = "free",
               labeller = labeller(group = label_wrap_gen(width = 1))) +
    xlab("Proportion") + ylab(NULL) +
    theme_custom() +
    scale_fill_manual(values = cols.paired, labels = mean.names) +
    scale_y_discrete(labels = whisker_to_sfcoy) +
    theme(axis.text.y = element_text(size = 4),
          axis.ticks.y = element_blank(),
          strip.text.y = element_text(size = 4, angle = 0, hjust = 0.5, vjust = 0.5),
          strip.background.y = element_blank(),
          legend.title = element_blank(),
          legend.text = element_text(size = 7),
          legend.key.size = unit(0.5, "cm"),
          legend.position = "bottom") +
    guides(fill = guide_legend(nrow = 1))

fig.4b
```


# Model with continuous effect and three dietary sources


## Prepare source data

```{r}
# Compute grand mean and pooled SD for each source
human_food_stats <- source_df %>% 
  filter(species %in% c("ff_chicken", "wet.food")) %>% 
  summarise(
    species = "human",
    Meand13C = mean(Meand13C),  
    Meand15N = mean(Meand15N),

    # Pooled standard deviation formula
    SDd13C = sqrt(sum((SDd13C^2 * (n - 1))) / sum(n - 1)),
    SDd15N = sqrt(sum((SDd15N^2 * (n - 1))) / sum(n - 1)),
    n = sum(n))

natural.low.N_stats <- source_df %>% 
  filter(species %in% c("fruit", "gopher")) %>% 
  summarise(
    species = "natural.low.N",
    Meand13C = mean(Meand13C),  
    Meand15N = mean(Meand15N),

    # Pooled standard deviation formula
    SDd13C = sqrt(sum((SDd13C^2 * (n - 1))) / sum(n - 1)),
    SDd15N = sqrt(sum((SDd15N^2 * (n - 1))) / sum(n - 1)),
    n = sum(n))

natural.high.N_stats <- source_df %>% 
  filter(species %in% c("medium.mammal", "rat")) %>% 
  summarise(
    species = "natural.high.N",
    Meand13C = mean(Meand13C),  
    Meand15N = mean(Meand15N),

    # Pooled standard deviation formula
    SDd13C = sqrt(sum((SDd13C^2 * (n - 1))) / sum(n - 1)),
    SDd15N = sqrt(sum((SDd15N^2 * (n - 1))) / sum(n - 1)),
    n = sum(n))

# Combine
three_source_df <- bind_rows(human_food_stats, natural.high.N_stats, natural.low.N_stats)

# Write to file
# write.csv(three_source_df, "Cleaned_Data/coyote_source_three.csv", row.names = FALSE)
```

## Prepare TDF data frame

```{r}
tdf_df_three <- data.frame(
  species = three_source_df$species,
  Meand13C = 3.97,
  SDd13C = 1.85,
  Meand15N = 2.76,
  SDd15N = 1.41)

# Write to csv
# write.csv(tdf_df_three, "Cleaned_Data/coyote_discrimination_three.csv", row.names = FALSE)
```

## Prepare consumer data

```{r}
# Load consumer data 
data_averages_isa <- read.csv("Cleaned_Data/averaged_whiskerdata.csv") %>% 
  filter(urban == "urban") %>%  # keep only SF coyotes
  filter(!is.na(group)) %>%  # remove individuals without known territories
  left_join(covs, by="group") # add covariate data

# Make consumer df
consumer_df_isa <- data.frame(
  d13C = data_averages_isa$C_avg,
  d15N = data_averages_isa$N_avg, 
  ISA = data_averages_isa$ISA) %>% 
  mutate_if(is.character, as.factor) 

# Write to file
# write.csv(consumer_df_isa, "Cleaned_Data/coyote_consumer_isa.csv", row.names = FALSE)
```

## Investigate mixing polygon

```{r}
# Make data frame for plotting with source values corrected by TDF factors
three_source_df_plotting <- three_source_df %>% 
  mutate(Meand13C = Meand13C + TDF.C,
         Meand15N = Meand15N + TDF.N)
  
# Plot prey sources
ggplot()+
  geom_point(data=three_source_df_plotting, mapping=aes(x=Meand13C, y=Meand15N, color=species), size=4)+
  geom_errorbar(data=three_source_df_plotting, mapping=aes(x=Meand13C, y=Meand15N,
                                                 ymax=Meand15N+SDd15N, ymin=Meand15N-SDd15N,
                                                 color=species, width=0.3))+
  geom_errorbar(data=three_source_df_plotting, mapping=aes(x=Meand13C, y=Meand15N,
                                                 xmax=Meand13C+SDd13C, xmin=Meand13C-SDd13C,
                                                 color=species, width=0.3))+
  geom_text(data=three_source_df_plotting, mapping=aes(x=Meand13C, y=Meand15N, label = species), vjust = -1)+
  geom_point(data=consumer_df_isa, mapping=aes(x=d13C, y=d15N))+
  theme_custom()+
  scale_y_continuous(limits=c(4,13), breaks = seq(4, 15, by = 1))+
  scale_x_continuous(limits=c(-25,-12), breaks = seq(-25, -10, by=1))+
  xlab(expression(delta^{13}~C)) +
  ylab(expression(delta^{15}~N))+
  theme(legend.position = "none")
```

## Create MixSIAR model objects

```{r}
# model with impervious surface cover
mix.coy.ISA <- load_mix_data(filename="Cleaned_Data/coyote_consumer_isa.csv", 
                     iso_names=c("d13C","d15N"), 
                     factors=NULL, 
                     fac_random=NULL, 
                     fac_nested=NULL, 
                     cont_effects="ISA")

# null model for comparison
mix.coy.ISA.null <- load_mix_data(filename="Cleaned_Data/coyote_consumer_isa.csv", 
                     iso_names=c("d13C","d15N"), 
                     factors=NULL, 
                     fac_random=NULL, 
                     fac_nested=NULL, 
                     cont_effects=NULL)

source.coy.ISA <- load_source_data(filename="Cleaned_Data/coyote_source_three.csv",
                           source_factors=NULL, 
                           conc_dep=FALSE, 
                           data_type="means", 
                           mix.coy.ISA)

discr.coy.ISA <- load_discr_data(filename="Cleaned_Data/coyote_discrimination_three.csv", mix.coy.ISA)
```

## Set prior:

```{r}
ISA.alpha <- c(43,17,40)
ISA.alpha <- ISA.alpha*length(ISA.alpha)/sum(ISA.alpha)

plot_prior(alpha.prior=ISA.alpha,
           source=source.coy.three,
           plot_save_pdf=FALSE,
           plot_save_png=FALSE,
           filename="prior_plot_ISA")
```

## Run model:

Write the JAGS model file and run it. This takes a long time, so it is recommended to run this model on an HPC cluster.

```{r}
# Write the JAGS model file
model_filename <- "MixSIAR_model_ISA.txt"
resid_err <- TRUE
process_err <- TRUE
write_JAGS_model(model_filename, resid_err, process_err, mix.coy.ISA, source.coy.ISA)

# Run full model
jags.ISA <- run_model(run="long", mix.coy.ISA, source.coy.ISA, discr.coy.ISA, model_filename, alpha.prior=ISA.alpha)

# Write the JAGS model file for the null model
model_filename <- "MixSIAR_model_ISA_null.txt"
resid_err <- TRUE
process_err <- TRUE
write_JAGS_model(model_filename, resid_err, process_err, mix.coy.ISA.null, source.coy.ISA)

# Run null model
jags.ISA.null <- run_model(run="long", mix.coy.ISA.null, source.coy.ISA, discr.coy.ISA, model_filename, alpha.prior=ISA.alpha)

# Load model output if not running the model
jags.ISA <- readRDS("Model_Output/jags.ISA.rds")
jags.ISA.null <- readRDS("Model_Output/jags.ISA.null.rds")
 
# Compare models
comparison.table <- compare_models(list(jags.ISA, jags.ISA.null))
comparison.table # model with impervious cover is much better fit

# Assess convergence with Gelman-Rubin and Geweke dignostics
output_diagnostics(jags.ISA, mix.coy.ISA, source.coy.ISA, output_options)

# Mean posterior estimates from model
df.stats.ISA <- output_stats(jags.ISA, mix.coy.ISA, source.coy.ISA, output_options)

# Extract posterior and turn into data frame
g.post.ISA <- output_posteriors(jags.ISA, mix.coy.ISA, source.coy.ISA, output_options)
posterior_cont <- as.data.frame((g.post.ISA[["cont"]][[1]][["data"]]))

# Plot
fig.4c <- ggplot() +
  geom_smooth(posterior_cont, mapping=aes(x=x, y=median, color=source))+
  geom_ribbon(posterior_cont, mapping=aes(x=x, ymin=low, ymax=high, fill=source), alpha=0.3)+
  labs(x="Impervious Surface Cover", y="Proportion")+
  theme_custom()+
  scale_y_continuous(limits=c(0,1), expand=c(0,0))+
  scale_x_continuous(limits=c(18.5,81), expand=c(0,0))+
  scale_color_manual(name=NULL,
                     values=c("#c0362d", "#2D2926", "#284A45"),
                     labels=c("Anthropogenic \nFood", "High-Trophic \nFood", "Low-Trophic \nFood"))+
  scale_fill_manual(name=NULL,
                    values=c("#db8c87", "#8C8076","#65948d" ),
                    labels=c("Anthropogenic \nFood", "High-Trophic \nFood", "Low-Trophic \nFood"),
                    guide = guide_legend(override.aes = list(alpha = 1)))+
  theme(legend.position = "bottom",
        legend.text = element_text(size=7),legend.key.size = unit(0.5, "cm"))

fig.4c
```

# Arrange figure:

```{r}
# Combine 4a and 4c vertically first
left_column <- ggarrange(fig.4a, fig.4c,
                         ncol = 1, 
                         nrow = 2,
                         align = "v")

# now add figure 4b
fig.4 <- ggarrange(left_column, fig.4b,
                   ncol = 2, 
                   widths = c(1, 1.5))

fig.4

# ggsave("Figures/Figure4.png", dpi=600,height=6, width=8.2)
```


